---
title: "Correlation and Causation"
subtitle: "What are they good for?"
author: "Carolina Torreblanca"
institute: "University of Pennsylvania"

format:
  revealjs:
    toc: false
    theme: [custom_iea.scss]
    width: 1050
    margin: 0.1
    logo: DevLab_Logo_29Mar2023.png
    footer: "https://carolina-torreblanca.github.io/psci3200-globaldev-main/"
    embed-resources: true
    template-partials:
      - title-slide.html
    highlight-style: atom-one-dark
    code-overflow: wrap

execute:
  eval: false
  echo: true
  warning: false
  message: false

editor: source
---

# Logistics

## Assignments

- Did everyone find the readings and slides for today?
- For next week:
  - Remember you have a quasi-assignment

---

# Correlation and Causation

## What are we trying to learn from data?

This lecture is not mainly about techniques.

It is about a conceptual problem that underlies all empirical social science:

How can we learn about cause and effect from data?

---

## Start with something familiar: correlation

What does it mean to say that two variables, X and Y, are correlated?

- They tend to move together
- When X is high, Y tends to be high (or low)
- There is a systematic pattern in observed data

---

## What correlation is

Correlation is a statement about patterns in observed data.

- It summarizes co-variation
- It is descriptive
- It does not, by itself, explain why the pattern exists

This feels straightforward.

---

## Now a harder question: causation

What does it mean to say that X causes Y?

Take a moment to think about this.

---

## Common intuitions about causation

People often answer:

- X leads to Y
- X produces Y
- X makes Y happen
- Y depends on X

These answers feel intuitive.

But notice what they do not tell us.

---

## The hidden problem

None of these answers explain:

How would we know that X causes Y?

They describe a relationship, but not a method for establishing it.

This is where the difficulty begins.

---

## A very old problem (Hume, simplified)

When we say “X causes Y,” we are really asking two questions:

- What do we mean by a cause?
- How could we ever be sure that X really causes Y?

Philosophers like David Hume pointed out that this is not obvious.

---

## Why causation is not observable

We never directly observe causation.

What we observe:
- Events
- Sequences
- Regularities

Example:
- You flip a switch
- The light turns on

You observe the switch move.
You observe the light turn on.

You do not observe causation itself.

---

## Why repetition does not solve the problem

Even if the switch always turns on the light:

- You still only observe regularity
- Not necessity
- Not causation itself

All the data in the world look the same whether causation is really there or not.

---

## Why this is a problem for science

If causation is not directly observable, then:

- Correlation alone cannot answer causal questions
- Observed patterns admit multiple explanations
- Data do not come labeled with causes

So how does science proceed?

---

## Different kinds of questions

It helps to distinguish different goals.

- Descriptive questions: What is happening?
- Predictive questions: Given X, what is likely to happen to Y?
- Causal questions: What would happen if we intervened?

These are not the same.

---

## Same data, different questions

Example:

People who live longer tend to take more vitamins.

- Descriptive: Do vitamin users live longer?
- Predictive: Can vitamin use help predict lifespan?
- Causal: Would people live longer if they started taking vitamins?

Only the last question is causal.

---

## The core difficulty

Causal questions require a comparison:

What actually happened  
versus  
what would have happened otherwise

But we never observe both.

This is the fundamental difficulty of causality.

---

# The Scientific Response

## A redefinition of causality

Rather than trying to observe causation directly, scientists made a practical move.

They defined causality in a way that could be studied.

This move came from applied problems, not philosophy.

---

## Historical origins

The modern approach to causality emerged from:
- Agriculture
- Field experiments
- Policy evaluation

For example:
What would crop yield be with fertilizer?
What would it be without fertilizer?

You cannot observe both on the same plot of land.

---

## The Neyman–Rubin contribution

Jerzy Neyman and Donald Rubin formalized this approach.

Their contribution is the **Potential Outcomes Framework**.

Causality is defined through comparisons between potential outcomes.

---

## Potential outcomes

For each unit \( i \), define:

- \( Y_i(1) \): outcome if unit \( i \) receives treatment
- \( Y_i(0) \): outcome if unit \( i \) does not receive treatment

These are potential outcomes.

---

## Defining causal effects

The individual-level causal effect is defined as:

\[
TE_i = Y_i(1) - Y_i(0)
\]

This is a definition: it tells us what we mean by a causal effect.

---

## The Fundamental Problem of Causal Inference (explicit)

For any unit \( i \), we can observe:

\[
Y_i = 
\begin{cases}
Y_i(1) & \\text{if } D_i = 1 \\\\
Y_i(0) & \\text{if } D_i = 0
\end{cases}
\]

But we can never observe both \( Y_i(1) \) and \( Y_i(0) \) for the same unit at the same time.

The unobserved potential outcome is counterfactual.

---

## Why this problem is fundamental

This is not a data limitation.

It is a logical one.

No amount of additional data lets us observe both potential outcomes for the same unit at the same time.

Causal inference is the craft of building credible counterfactual comparisons.

## Correlations: Necessary Components

\

**What do we need to calculate correlations?**

- Measures of central tendency
  + Mean 
- Measures of spread
  + Variance
  + Standard deviation


## Central Tendency: Mean


$$
\mu_X = \frac{1}{n} \sum_{i}^{n} X_i
$$

```{r}
#| echo: true

my_vector = rnorm(10, mean = 10, sd = 5)
# Step 1: Sum the values
sum_values <- sum(my_vector)
# Step 2: Count the number of elements
count_elements <- length(my_vector)
# Step 3: Calculate the mean
mean_value <- sum_values / count_elements

print(mean_value)
mean(my_vector)
```


## Spread: Variance

$$
\sigma^2_X = \frac{1}{N} \sum_{i}^{N} (X_i - \mu_X)^2
$$

::: {.incremental}
- What does the square in $\sigma^2$ accomplish?
- What are the implications for interpretation?
  + Units
  + Distribution
- Even with these basic measures, we're already thinking about the distribution!

:::

## Spread: Variance

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Create a vector"

## Create vector, sort by size, and store var
set.seed(123)
dat = rnorm(10, mean = 10, sd = 5)
dat = sort(dat)
o_var = var(dat)
print(dat)

```

```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Add a constant to a big number"

## Create new dataframe for big addition and store vector length
b_dat = dat
ind = length(b_dat)

## Add four to the largest number in the vector and calculate size of var increase
b_dat[ind] = b_dat[ind] + 4
b_var = var(b_dat)
val = b_var - o_var
cat("Variance increases by", val )
```



```{r}
#| echo: true
#| code-fold: true
#| code-summary: "Add a constant to a smaller number"

## Create new dataframe for small addition
s_dat = dat

## Add four to the smallest number in the vector and calculate size of var increase
s_dat[ind-2] = s_dat[ind-2] + 4
s_var = var(s_dat)
val = s_var - o_var
cat("Variance increases by", val )

```

## Spread: Standard Deviation

$$
\sigma_X = \sqrt{\frac{1}{N} \sum_{i}^{N} (X_i - \mu_X)^2}
$$

::: {.incremental}
- What does the $\sqrt{}$ accomplish?
- What are the implications for interpretation?
  + Expressed in the same units as the observations
  + How far we expect each observation to be from the mean, on average
- This means we can report effect sizes as SDs
:::


## Measures of Correlation

::: {.incremental}
- <span style="color:red;">Covariance</span> $\text{Cov}(X, Y) = \frac{1}{n} \sum_{i}^{N} (X_i - \bar{X})(Y_i - \bar{Y})$
  + Product of the deviations
  + Range: unbounded
- <span style="color:red;">Correlation coefficient</span> $\text{Cor}(X, Y) = \frac{\text{Cov}(X, Y)}{\sigma_X \sigma_Y}$
  + Covariance normalized by product of SDs
  + Range: -1 to 1
- <span style="color:red;">Slope</span> $\beta_X = \frac{\text{Cov}(X, Y)}{\sigma^2_X}$
  + Covariance normalized by variance
  + Expected change in $Y$ with 1-unit change in $X$
:::


## Measures of Correlation

::: {.incremental}
- What does the correlation coefficient tell you that slope doesn't?
  + Consistency of the relationship on bounded scale (-1 to 1)
- What does slope tell you that the correlation coefficient doesn't?
  + Substantive importance (magnitude)
- Give an example of when you'd prefer each
  + Correlation: When comparing relationships on different scales
  + Slope: When thinking about ROI
:::


## Correlation

::: columns
::: {.column width="70%"}

\
**What can with do with them?**

- <span style="color:red;">Description</span>: quantitative comparisons 
:::

::: {.column width="30%"}

::: {.fragment}

::: {.callout-note appearance="simple" icon=false} 
-   <span style="color:red;">*sample matters alot*</span>
-   <span style="color:blue;">*sample matters less*</span>
:::

:::

:::
:::
- <span style="color:red;">Forecasting</span>: sample population $\rightarrow$ out-of-sample 
- <span style="color:blue;">Causal inference</span>: correlation + research design

::: {.fragment}

**Simple, but powerful**

- Non-linearities, interactions, machine learning

:::


# Causation

## Schools of Thought

- Potential outcomes and counterfactuals (Econ)
- DAGs and do-calculus (CS)
- Manipulability (Philosophy)


“We think of a cause as something that makes a difference, and the difference it makes must be a difference from what would have happened without it.” (Lewis, 1973)


## Causality: Why bother?

- Understanding cause and effect is how we change things in the real world
- Causal inference separates good evaluations from bad
  + Policy change
  + Development intervention
- Causal identification is not binary
  + It's harder for some policies and interventions than others
  + Variety of tools that can help us rule out different threats to inference

## Causality: Why bother?

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"
library(ggplot2)

Year = c(0,1,2,3)
Outcome = c(NA, 1.2, 1.4,NA)
Treatment = c("Control", "Control","Control","Control")

dat = data.frame(Year, Outcome, Treatment)

ggplot(data = dat, aes(x = Year, y = Outcome, group = Treatment)) +
  geom_line(aes(linetype=Treatment),size=2) +
  geom_point(size = 6) +
  scale_linetype_manual(values=c("solid")) +
  xlim(0,3) + 
  scale_y_continuous(limits = c(1,1.85), breaks = seq(1, 1.85, by = .1)) + 
  theme(legend.position = "none", text = element_text(size=20)) 
 
```


## Causality: Why bother?


```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

Year = c(0,1,2,3)
Outcome = c(1, 1.2, 1.4, 1.6, 
            0.9, 1.3, 1.7, 2.1)
Treatment = c("Control", "Control","Control","Control", 
              "Treatment", "Treatment", "Treatment", "Treatment")

dat = data.frame(Year, Outcome, Treatment)


ggplot(data = dat, aes(x = Year, y = Outcome, group = Treatment)) +
  geom_line(aes(linetype=Treatment),size=2) +
  geom_point(size = 6) +
  xlim(0,3) + 
  scale_y_continuous(breaks = seq(1, 1.85, by = .1)) + 
  scale_linetype_manual(values=c("solid", "solid")) +
  coord_cartesian(ylim = c(1, 1.85), clip = "on") +
  theme(legend.position = "none", text = element_text(size=20)) 

```

## Causality: Why bother?

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

Year = c(0,1,2,3)
Outcome = c(1, 1.2, 1.4,1.6)
Treatment = c("Control", "Control","Control","Control")

dat = data.frame(Year, Outcome, Treatment)

ggplot(data = dat, aes(x = Year, y = Outcome, group = Treatment)) +
  geom_line(aes(linetype=Treatment),size=2) +
  geom_point(size = 6) +
  scale_linetype_manual(values=c("solid")) +
  xlim(0,3) + 
  scale_y_continuous(limits = c(1,1.85), breaks = seq(1, 1.85, by = .1)) + 
  theme(legend.position = "none", text = element_text(size=20))

```

## Causality: Why bother?

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

Year = c(0,1,2,3)
Outcome = c(NA, 1.2, 1.4, NA, 
            NA, 1.3, 1.7, NA, 
            NA, 1.3, 1.5, NA)
Treatment = c("Control", "Control","Control","Control", 
              "Treatment", "Treatment", "Treatment", "Treatment",
              "Comparison","Comparison","Comparison","Comparison")

dat = data.frame(Year, Outcome, Treatment)


ggplot(data = dat, aes(x = Year, y = Outcome, group = Treatment)) +
  geom_line(aes(linetype=Treatment),size=2) +
  geom_point(size = 6) +
  xlim(0,3) + 
  scale_y_continuous(limits = c(1,1.85), breaks = seq(1, 1.85, by = .1)) + 
  scale_linetype_manual(values=c("dotted", "solid", "solid")) +
  theme(legend.position = "none", text = element_text(size=20)) 

```



## Causality: Why bother?

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

Year = c(0,1,2,3)
Outcome = c(NA, 1.2, 1.4, NA, 
            NA, 1.3, 1.7, NA)
Treatment = c("Control", "Control","Control","Control", 
              "Treatment", "Treatment", "Treatment", "Treatment")

dat = data.frame(Year, Outcome, Treatment)


ggplot(data = dat, aes(x = Year, y = Outcome, group = Treatment)) +
  geom_line(aes(linetype=Treatment),size=2) +
  geom_point(size = 6) +
  xlim(0,3) + 
  scale_y_continuous(limits = c(1,1.85), breaks = seq(1, 1.85, by = .1)) + 
  scale_linetype_manual(values=c("solid", "solid")) +
  theme(legend.position = "none", text = element_text(size=20))

```



## Causality: Why bother?

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

Year = c(0,1,2,3)
Outcome = c(1, 1.2, 1.4,1.6, 
            1.1, 1.3, 1.7, 1.9, 
            1.1, 1.3, 1.5, 1.7)
Treatment = c("Control", "Control","Control","Control", 
              "Treatment", "Treatment", "Treatment", "Treatment",
              "Comparison","Comparison","Comparison","Comparison")

dat = data.frame(Year, Outcome, Treatment)


ggplot(data = dat, aes(x = Year, y = Outcome, group = Treatment)) +
  geom_line(aes(linetype=Treatment),size=2) +
  geom_point(size = 6) +
  scale_y_continuous(breaks = seq(1, 1.9, by = .1)) + 
  scale_linetype_manual(values=c("dotted", "solid", "solid")) +
  coord_cartesian(ylim = c(1, 1.85), clip = "on") +
  theme(legend.position = "none", text = element_text(size=20))

```



## Causality: Why bother?

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

Year = c(0,1,2,3)
Outcome = c(1, 1.2, 1.4,1.6, 
            1.1, 1.3, 1.7, 1.9)
Treatment = c("Control", "Control","Control","Control", 
              "Treatment", "Treatment", "Treatment", "Treatment")

dat = data.frame(Year, Outcome, Treatment)


ggplot(data = dat, aes(x = Year, y = Outcome, group = Treatment)) +
  geom_line(aes(linetype=Treatment),size=2) +
  geom_point(size = 6) +
  scale_y_continuous(breaks = seq(1, 1.9, by = .1)) + 
  scale_linetype_manual(values=c("solid", "solid")) +
  coord_cartesian(ylim = c(1, 1.85), clip = "on") +
  theme(legend.position = "none", text = element_text(size=20))

```



## Causality: What makes it hard?

::: {.fragment}

**Fundamental Problem of Causal Inference**

$$
Y_i = 
\begin{cases} 
Y_i(1) & \text{if } D_i = 1 \text{ (treatment group)} \\
Y_i(0) & \text{if } D_i = 0 \text{ (control group)}
\end{cases}
$$

:::

::: {.fragment}

- We only observe any given unit in one treatment status at any one time so we can never directly observe the causal effect of a treatment on a unit. 

:::

## Potential Outcomes and Counterfactuals

::: {.fragment}

**Treatment Effect for individual $i$**

- $TE_i = Y_i(1) - Y_i(0)$

:::

::: {.fragment}

**Average Treatment Effect (ATE)**

- $ATE = \frac{1}{N} \sum_{i=1}^{N} TE_i$

:::

## Many Different Tools

- Randomized experiments
  + Gold-standard
  + Field and survey
- Observational data
  + Natural experiments
  + Difference-in-Differences
  + Matching, Synthetic Control
<!-- - You always need to think about the outcomes and measurement -->
<!--   + Behavioral vs attitudinal -->
<!--   + Observed vs self-reported -->

## DAGs and Confounding

```{r}
library(ggdag)

dagify(
  x ~ a,
  y ~ x,
  y ~ a

) %>% ggdag()

```


## Next Meeting

- Randomized experiments
  + Review of FPCI
  + How randomization addresses confounding
- The role of social science researchers?
  + What should people with social science training be doing with their skills?
- Final Project
  + What are your options?
  + Possible data sources

<!-- **The correlation between two variables is “the extent to which they tend to occur together” (p. 14). -->
<!-- A relation between two variables (or “features of the world”). -->
<!-- Two variables/features are: -->
<!-- Positively correlated if the two tend to occur together -->
<!-- Negatively correlated if one feature occurs when the other does not -->
<!-- Uncorrelated if the occurrence of one does not affect the occurrence of the other. -->

<!-- ## Politicians facing scandals tend to win re-election -->

<!-- - what data does this rely on? -->
<!-- - why is this not a correlation: we don't learn if these things tend to happen together -->
<!-- - what would make it a correlation  -->
<!-- - what data would we need to assess: rate of re-election for both scandal and non-scandal politicians -->

<!-- ::: columns -->
<!-- ::: {.column width="50%"} -->
<!-- ::: -->

<!-- ::: {.column width="50%"} -->
<!-- ::: -->
<!-- ::: -->
