---
title: "Introduction to OLS"
subtitle: "Linear Regression"
author: "Carolina Torreblanca"
institute: "University of Pennsylvania"

format:
  revealjs:
    toc: false
    theme: [custom_iea.scss]
    width: 1050
    margin: 0.1
    logo: DevLab_Logo_29Mar2023.png
    footer: "https://carolina-torreblanca.github.io/psci3200-globaldev-main/"
    embed-resources: true
    template-partials:
    - title-slide.html
    gfm:
    mermaid-format: png
editor: source
---

```{r}
#| include: false
options(scipen = 999)
```


## Agenda

\

1. Why OLS?
2. Two things regression does
3. From difference-in-means to regression
4. Continuous predictors and the regression line
5. Playing God: simulating data
6. Adding controls
7. Running and interpreting OLS in R


# Why OLS?

## The Workhorse of Social Science

\

::: {.fragment}
Almost every empirical paper you've read in this course uses **linear regression**.
:::

::: {.fragment}
::: {.incremental}
- Comparing a treatment group to a control group? **Regression.**
- Controlling for confounders? **Regression.**
- Fixed effects? **Regression** with extra variables.
- Difference-in-Differences? **Regression** with an interaction term.
- Your final project? **Regression.**
:::
:::

## Why Should You Care?

\

::: {.fragment}
For your final projects, you need to:
:::

::: {.fragment}
::: {.incremental}
1. Estimate a relationship between variables
2. Control for things that might confound that relationship
3. Present and interpret your results
4. All of this is regression
:::
:::

::: {.fragment}
\
Today: what regression actually does, and how to use it.
:::


# Two Things Regression Does

## Thing 1: Best Guess of Y

\

::: {.fragment}
**"Given what I know about X, what is my best guess for Y?"**
:::

::: {.fragment}
::: {.incremental}
- If I tell you a country's GDP per capita is $10,000...
- What's your best guess for its life expectancy?
- Regression gives you that guess — and a formula for making it
:::
:::

## Thing 2: How Much Does Y Change?

\

::: {.fragment}
**"When X goes up by one unit, how much does Y change?"**
:::

::: {.fragment}
::: {.incremental}
- If GDP per capita increases by $1,000...
- How many more years of life expectancy do we predict?
- This is the **slope** — one number that summarizes the relationship
:::
:::

::: {.fragment}
\
These are two sides of the same coin. The line gives you both.
:::


# From Difference-in-Means to Regression

## Something You Already Know

\

Comparing a treatment group to a control group:

$$
\text{Effect} = \overline{Y}_{treatment} - \overline{Y}_{control}
$$

::: {.fragment}
You've been doing this since the causality lectures.

**This is regression.**
:::


## Let's Prove It

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"
library(ggplot2)
library(modelsummary)

# Simulate an experiment
set.seed(42)
experiment <- data.frame(
  treatment = c(rep(0, 50), rep(1, 50))
)
experiment$outcome <- 10 + 3 * experiment$treatment + rnorm(100, 0, 2)
```

::: {.fragment}
**Method 1: Difference in means**
```{r}
#| echo: true
mean(experiment$outcome[experiment$treatment == 1]) -
  mean(experiment$outcome[experiment$treatment == 0])
```
:::

::: {.fragment}
**Method 2: Regression**
```{r}
#| echo: true
coef(lm(outcome ~ treatment, data = experiment))
```
:::

## Same Answer

\

::: {.incremental}
- The intercept = the control group mean
- The slope on `treatment` = the difference in means
- **Regression with a binary variable IS a difference in means**
- So you already understand the core idea — now we generalize it
:::


## Why Not Just Use Difference-in-Means for Everything?

\

::: {.fragment}
Because the real world is more complicated:
:::

::: {.fragment}
::: {.incremental}
- What if your key variable isn't binary? (GDP is continuous)
- What if you need to **control** for something? (Education, geography...)
- What if you want to compare across **multiple groups** at once?
- Regression handles all of this. Difference-in-means can't.
:::
:::


# The Regression Line

## Now With a Continuous Variable

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

set.seed(42)
n <- 80
countries <- data.frame(
  gdp = runif(n, 1, 50)  # GDP per capita in thousands of USD
)
countries$life_exp <- 55 + 0.4 * countries$gdp + rnorm(n, 0, 4)

ggplot(countries, aes(x = gdp, y = life_exp)) +
  geom_point(size = 3, alpha = 0.6, color = "#011F5B") +
  labs(x = "GDP per capita (thousands USD)",
       y = "Life expectancy (years)",
       title = "Each dot is a country") +
  theme_minimal(base_size = 18)
```

::: {.fragment}
There's clearly a pattern. But what line best captures it?
:::

## Which Line?

```{r}
#| echo: false
#| warning: false

ggplot(countries, aes(x = gdp, y = life_exp)) +
  geom_point(size = 3, alpha = 0.6, color = "#011F5B") +
  geom_abline(intercept = 50, slope = 0.6, color = "red",
              linewidth = 1.2, alpha = 0.7) +
  geom_abline(intercept = 60, slope = 0.2, color = "orange",
              linewidth = 1.2, alpha = 0.7) +
  geom_abline(intercept = 45, slope = 0.5, color = "purple",
              linewidth = 1.2, alpha = 0.7) +
  labs(x = "GDP per capita (thousands USD)",
       y = "Life expectancy (years)",
       title = "Three candidate lines — which is best?") +
  theme_minimal(base_size = 18)
```

::: {.fragment}
We need a rule for "best." That rule is **OLS**.
:::


## The Equation

\

A line needs two numbers — an **intercept** and a **slope**:

$$
Y_i = \alpha + \beta X_i + \epsilon_i
$$

::: {.fragment}
::: {.incremental}
- $\alpha$ (**intercept**): our best guess of $Y$ when $X = 0$
- $\beta$ (**slope**): how much $Y$ changes when $X$ goes up by one unit
:::
:::

## What About $\epsilon$?

\

$$
Y_i = \alpha + \beta X_i + \epsilon_i
$$

::: {.fragment}
Life expectancy depends on a million things: healthcare, diet, war, genetics, luck...
:::

::: {.fragment}
::: {.incremental}
- We're only using GDP to predict life expectancy
- Everything else lands in $\epsilon_i$ (the **error term**)
- The error is not a mistake — it's the **gap between our simple model and complicated reality**
:::
:::


## How OLS Picks the Line

\

For each data point, the line makes a **prediction**:

$$
\text{Prediction for country } i = \alpha + \beta X_i
$$

::: {.fragment}
The **residual** is how far off that prediction is:

$$
\text{Residual}_i = \text{Observed } Y_i - \text{Prediction}_i
$$
:::

## Minimizing Residuals

\

**OLS** picks the $\alpha$ and $\beta$ that make the residuals as small as possible.

::: {.fragment}
Specifically, it minimizes the **sum of squared residuals**:

$$
SSR = \sum_{i=1}^{N} \text{Residual}_i^2
$$
:::

::: {.fragment}
::: {.incremental}
- Why squared? So positive and negative misses don't cancel out
- And big misses are penalized more than small ones
:::
:::

## Visualizing Residuals

```{r}
#| echo: false
#| warning: false

model <- lm(life_exp ~ gdp, data = countries)
countries$predicted <- predict(model)

set.seed(1)
idx <- sample(1:n, 20)
sub <- countries[idx, ]

ggplot(countries, aes(x = gdp, y = life_exp)) +
  geom_point(size = 3, alpha = 0.4, color = "#011F5B") +
  geom_smooth(method = "lm", se = FALSE, color = "#011F5B", linewidth = 1.2) +
  geom_segment(data = sub,
               aes(x = gdp, xend = gdp, y = life_exp, yend = predicted),
               color = "red", linewidth = 0.8, alpha = 0.7) +
  labs(x = "GDP per capita (thousands USD)",
       y = "Life expectancy (years)",
       title = "Red segments = residuals",
       subtitle = "The gap between what we observed and what the line predicts") +
  theme_minimal(base_size = 18)
```

::: {.fragment}
OLS = the line that makes these red segments as small as possible (in total).
:::


# Playing God

## A Simulation

\

::: {.fragment}
Imagine you are God. You get to decide how the world works.
:::

::: {.fragment}
You decree: **life expectancy = 55 + 0.4 $\times$ GDP**
:::

::: {.fragment}
That's the general rule. But you're a generous God — each country gets a little bit of **free will** (randomness).
:::

::: {.fragment}
\
**Question:** If a researcher only sees the data — not your rule — can they figure out what you decided?
:::

## You Create the World

```{r}
#| echo: true
#| warning: false

set.seed(123)

# GOD'S RULE (the truth no one can see)
true_intercept <- 55
true_slope     <- 0.4

# Generate data for 100 countries
n <- 100
gdp <- runif(n, 1, 50)  # GDP in thousands of USD
life_exp <- true_intercept + true_slope * gdp + rnorm(n, 0, 4)

god_data <- data.frame(gdp, life_exp)
```

::: {.fragment}
\
Now let's pretend we're a researcher who **only sees the data**, not the rule.
:::

## OLS Tries to Recover the Truth

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

fit <- lm(life_exp ~ gdp, data = god_data)

ggplot(god_data, aes(x = gdp, y = life_exp)) +
  geom_point(alpha = 0.5, color = "#011F5B") +
  geom_abline(intercept = true_intercept, slope = true_slope,
              color = "red", linewidth = 1.5, linetype = "dashed") +
  geom_abline(intercept = coef(fit)[1], slope = coef(fit)[2],
              color = "#011F5B", linewidth = 1.5) +
  annotate("label", x = 8, y = 73, label = "God's truth",
           color = "red", size = 5) +
  annotate("label", x = 8, y = 70, label = "OLS estimate",
           color = "#011F5B", size = 5) +
  labs(x = "GDP per capita (thousands USD)", y = "Life expectancy",
       title = "OLS gets very close to the truth") +
  theme_minimal(base_size = 18)
```

## How Close?

```{r}
#| echo: true

# What God decided
c(true_intercept = true_intercept, true_slope = true_slope)

# What OLS found
coef(fit)
```

::: {.fragment}
\
Not perfect — because of the free will (randomness) — but very close.
:::

## Different Worlds, Same Rule

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

par_df <- data.frame()
for (world in 1:3) {
  set.seed(world * 10)
  g <- runif(n, 1, 50)
  le <- true_intercept + true_slope * g + rnorm(n, 0, 4)
  f <- lm(le ~ g)
  par_df <- rbind(par_df, data.frame(
    gdp = g, life_exp = le,
    world = paste("World", world,
                  "  (slope =", round(coef(f)[2], 2), ")")
  ))
}

ggplot(par_df, aes(x = gdp, y = life_exp)) +
  geom_point(alpha = 0.4, color = "#011F5B") +
  geom_smooth(method = "lm", se = FALSE, color = "#011F5B", linewidth = 1) +
  geom_abline(intercept = true_intercept, slope = true_slope,
              color = "red", linewidth = 1, linetype = "dashed") +
  facet_wrap(~world) +
  labs(x = "GDP per capita (thousands USD)", y = "Life expectancy",
       title = "Same rule, different randomness — OLS is close each time",
       subtitle = "Red dashed = God's truth (slope = 0.4)") +
  theme_minimal(base_size = 14)
```

## OLS Works... When the Truth Is Linear

\

::: {.incremental}
- We made a **linear** rule, and OLS (which fits a **line**) recovered it
- But what if the truth isn't linear?
- What if God decided something more complicated?
:::


## What If the Truth Is Curved?

```{r}
#| echo: true
#| warning: false
#| code-fold: true
#| code-summary: "Show code"

set.seed(42)
x_sim <- runif(200, 0, 10)

# Three different truths — with clear, big coefficients
linear_y    <- 10 + 3 * x_sim + rnorm(200, 0, 2)
quadratic_y <- 10 + 8 * x_sim - 0.8 * x_sim^2 + rnorm(200, 0, 2)
cubic_y     <- 5 + 12 * x_sim - 3 * x_sim^2 + 0.2 * x_sim^3 + rnorm(200, 0, 2)

sim_worlds <- rbind(
  data.frame(x = x_sim, y = linear_y,    truth = "Y = 10 + 3X  (linear)"),
  data.frame(x = x_sim, y = quadratic_y, truth = "Y = 10 + 8X - 0.8X²  (quadratic)"),
  data.frame(x = x_sim, y = cubic_y,     truth = "Y = 5 + 12X - 3X² + 0.2X³  (cubic)")
)

ggplot(sim_worlds, aes(x = x, y = y)) +
  geom_point(alpha = 0.3, color = "#011F5B", size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1.2) +
  geom_smooth(method = "loess", se = FALSE, color = "forestgreen",
              linewidth = 1.2, linetype = "dashed") +
  facet_wrap(~truth, scales = "free_y") +
  labs(x = "X", y = "Y",
       title = "Red = OLS line | Green dashed = actual pattern",
       subtitle = "OLS forces a straight line even when the truth is curved") +
  theme_minimal(base_size = 14)
```

## All Models Are Wrong, But Some Are Useful

\

::: {.incremental}
- OLS **always** fits a straight line
- When the truth is linear, it works great
- When the truth is curved, the line **misses the pattern**
- In practice, OLS is surprisingly robust unless things are dramatically nonlinear
- We mostly use OLS to **estimate associations and marginal effects**, not to predict
:::


# Adding Controls

## The Problem: Omitted Variables

\

::: {.fragment}
What if we left something important out of the model?
:::

::: {.fragment}
\
Let's play God again — but this time with **two** variables that matter.
:::

## God's New Rule

```{r}
#| echo: true
#| warning: false

set.seed(42)
n_c <- 80
ctrl <- data.frame(gdp = runif(n_c, 1, 50)) # GDP in thousands

# Education is correlated with GDP (richer countries → more education)
ctrl$education <- 2 + 0.4 * ctrl$gdp + rnorm(n_c, 0, 3)

# God's truth: BOTH matter, but education matters a LOT
ctrl$life_exp <- 45 + 0.2 * ctrl$gdp + 1.5 * ctrl$education + rnorm(n_c, 0, 2)
```

::: {.fragment}
\
God decided:

- GDP matters a little (true slope = **0.2**)
- Education matters a lot (true slope = **1.5**)
- But GDP and education are **correlated**
:::

## What Happens If We Forget Education?

```{r}
#| echo: false
#| warning: false

model_no_ctrl   <- lm(life_exp ~ gdp, data = ctrl)
model_with_ctrl <- lm(life_exp ~ gdp + education, data = ctrl)

modelsummary(list("GDP only" = model_no_ctrl,
                  "GDP + Education" = model_with_ctrl),
  estimate  = "{estimate}{stars} ({std.error})",
  statistic = NULL,
  gof_omit = 'IC|RMSE|Log|F|R2$|Std.')
```

## What Just Happened?

\

::: {.incremental}
- **Without education**: the GDP coefficient is inflated
- **With education**: it **drops dramatically**, closer to God's truth (0.2)
- Why? GDP and education are correlated — GDP was getting credit for education's work
- This is **omitted variable bias**
:::

## Omitted Variable Bias, Visually

```{r}
#| echo: false
#| warning: false
library(ggdag)

ovb_coords <- list(
  x = c(GDP = 1, LifeExp = 3, Education = 2),
  y = c(GDP = 1, LifeExp = 1, Education = 2)
)

dagify(
  LifeExp ~ GDP + Education,
  Education ~ GDP,
  exposure = "GDP",
  outcome = "LifeExp",
  coords = ovb_coords
) |>
  tidy_dagitty() |>
  node_status() |>
  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges(edge_width = 1,
    arrow_directed = grid::arrow(length = grid::unit(0.6, "lines"),
                                 type = "closed")) +
  geom_dag_point(aes(color = status), size = 22) +
  geom_dag_text(color = "black", size = 5) +
  scale_color_manual(values = c("#FF4136", "#0074D9"),
                     na.value = "#7FDBFF") +
  guides(color = "none") +
  theme_dag() +
  labs(title = "Education is a confounder",
       subtitle = "It affects both GDP and life expectancy")
```

::: {.fragment}
Without controlling for Education, the GDP $\rightarrow$ LifeExp path picks up the GDP $\rightarrow$ Education $\rightarrow$ LifeExp path too.
:::


## Interpretation With Controls

\

::: {.fragment}
**Without controls:**

"A $1,000 increase in GDP is associated with a `r round(coef(model_no_ctrl)[2], 2)` year increase in life expectancy."
:::

::: {.fragment}
**With controls:**

"Holding education constant, a $1,000 increase in GDP is associated with a `r round(coef(model_with_ctrl)[2], 2)` year increase."
:::

::: {.fragment}
\
"Holding constant" = comparing countries with the **same** education level.
:::


# Controls and Causality

## Does Controlling Make It Causal?

\

::: {.fragment}
Not automatically.
:::

::: {.fragment}
::: {.incremental}
- Remember the causality lectures: a causal effect is about the **counterfactual**
- "What **would have happened** to this country if its GDP were higher, but everything else stayed the same?"
- That's not a question about the best fit line — it's an **assumption** about the data generating process
:::
:::

## When Can We Make Causal Claims?

\

::: {.fragment}
We need a **good reason** for why the contrasts we're estimating are causal.
:::

::: {.fragment}
::: {.incremental}
- That reason always comes back to the same thing: the **average counterfactual is valid**
- The comparison group is a good stand-in for what would have happened without the treatment
- This is an assumption about the **world**, not about the model
:::
:::

## Where Do Good Reasons Come From?

\

::: {.fragment}
They rely on the **data generating process** — how the data came to be:
:::

::: {.fragment}
::: {.incremental}
- **Randomized experiments**: we controlled assignment, so treatment and control are comparable by design
- **Difference-in-Differences**: we use time variation and assume parallel trends
- **Regression Discontinuity**: we exploit a cutoff that creates quasi-random assignment
- **Instrumental Variables**: we find an exogenous source of variation
:::
:::

::: {.fragment}
\
All of these **use** regression to estimate effects — but the causal claim comes from the **design**, not the model.
:::


# Running OLS in R

## The Full Workflow

```{r}
#| echo: true
#| warning: false

# Fit the model
model <- lm(life_exp ~ gdp, data = countries)

# See the full output
summary(model)
```

## What Each Piece Means

\

::: {.incremental}
- **Estimate**: the intercept and slope — our best guesses of $\alpha$ and $\beta$
- **Std. Error**: how precise our estimate is (more on this in the uncertainty lecture)
- **Pr(>|t|)**: p-value — how likely we'd see an estimate this large if the true value were zero (also uncertainty lecture)
- **R-squared**: fraction of variation in $Y$ explained by $X$ (0 = nothing, 1 = everything)
:::

::: {.fragment}
\
For today, focus on the **Estimate** column.
:::

## Interpreting the Slope

```{r}
#| echo: true
coef(model)
```

::: {.fragment}
\
**Interpretation:** "A $1,000 increase in GDP per capita is associated with a `r round(coef(model)[2], 2)` year increase in life expectancy."
:::

::: {.fragment}
Or: "A **$10,000** increase in GDP is associated with a **`r round(coef(model)[2] * 10, 1)`** year increase in life expectancy."
:::


## Clean Output with `modelsummary()`

```{r}
#| echo: true
#| warning: false

modelsummary(model,
  estimate  = "{estimate}{stars} ({std.error})",
  statistic = NULL,
  gof_omit = 'IC|RMSE|Log|F|R2$|Std.')
```

## Visualize the Line

```{r}
#| echo: true
#| warning: false

ggplot(countries, aes(x = gdp, y = life_exp)) +
  geom_point(size = 3, alpha = 0.6, color = "#011F5B") +
  geom_smooth(method = "lm", color = "red", fill = "red", alpha = 0.15) +
  labs(x = "GDP per capita (thousands USD)", y = "Life expectancy",
       title = "OLS line with confidence band") +
  theme_minimal(base_size = 18)
```


# Wrapping Up

## What Regression Gives You

\

::: {.incremental}
1. A **best guess** of $Y$ for any value of $X$ (prediction)
2. A **slope** that tells you how $Y$ changes when $X$ changes (association)
3. The ability to **control** for other variables (reducing omitted variable bias)
4. A framework that everything else in this course builds on
:::

## What Regression Does NOT Give You

\

::: {.incremental}
- Causality does not come from the model — it comes from the **research design**
- The best fit line can be wrong if the relationship isn't linear — but in practice OLS is surprisingly robust
- Controls only help if you've identified the right confounders (think about your DAGs!)
:::
