---
title: "Data Wrangling Workshop: Olympics & Development"
author: "Carolina Torreblanca"

institute: "University of Pennsylvania"
date: last-modified
toc: true

format:
  html:
    self-contained: true

editor: source
---

# Part 0: Project Setup

Before we write any code, we need to set up a proper project structure. This is how you should organize **every** data analysis project, including your final project.

## 0.1 Create a project folder

On your computer, create a folder called `wrangling_workshop`. Inside it, create **three subfolders**:

```
wrangling_workshop/
├── input/
├── output/
├── code/
```

- `input/` — raw data goes here. You **never modify** files in this folder.
- `output/` — cleaned data, figures, and tables go here.
- `code/` — your `.qmd` or `.R` scripts go here.

**Do this now.** You can use Finder/File Explorer or the terminal:

```bash
mkdir wrangling_workshop
cd wrangling_workshop
mkdir input output code
```

## 0.2 Create an RProject

1. Open RStudio
2. Go to **File → New Project → Existing Directory**
3. Browse to your `wrangling_workshop` folder
4. Click **Create Project**

This creates a `.Rproj` file in your folder. From now on, **always open your project by double-clicking the `.Rproj` file**. This ensures that R knows where your files are.

**Why does this matter?** When you use an RProject, R sets the **working directory** to your project folder. So instead of writing:

```r
# Bad: absolute path that only works on YOUR computer
read_csv("/Users/carolina/Desktop/classes/psci3200/data/olympics_raw.csv")
```

You can write:

```r
# Good: relative path that works on anyone's computer
read_csv("input/olympics_raw.csv")
```

## 0.3 Download the data

Download these two CSV files and save them in your `input/` folder:

1. **Olympics data:** [olympics_raw.csv](https://github.com/carolina-torreblanca/psci3200-globaldev-main/raw/main/workshops/data_wrangling/olympics_raw.csv)
2. **World Bank indicators:** [wb_indicators_2016.csv](https://github.com/carolina-torreblanca/psci3200-globaldev-main/raw/main/workshops/data_wrangling/wb_indicators_2016.csv)

Your folder should now look like this:

```
wrangling_workshop/
├── wrangling_workshop.Rproj
├── input/
│   ├── olympics_raw.csv
│   └── wb_indicators_2016.csv
├── output/
├── code/
```

## 0.4 Create your script

1. In RStudio, go to **File → New File → Quarto Document**
2. Save it as `code/workshop.qmd`

Now paste this setup chunk at the top of your `.qmd` file and run it:

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(modelsummary)
```


# Part 1: Load and Explore the Olympics Data

This dataset contains every Olympic athlete and event from Athens 1896 to Rio 2016. Each row is one athlete in one event.

## 1.1 Read in the data

```{r}
#| warning: false
#| message: false

# Read the Olympics data from the input folder
olympics_raw <- read_csv("input/olympics_raw.csv")
```

Notice we used a **relative path** (`input/olympics_raw.csv`) not an absolute path. This works because our RProject set the working directory to the project folder.

## 1.2 Get to know your data

Before cleaning, always explore what you're working with.

```{r}
# How many rows and columns?
dim(olympics_raw)

# What are the variable names and types?
glimpse(olympics_raw)

# Look at the first few rows
head(olympics_raw)
```

**What do we notice?**

- Each row is an athlete-event (not a country or a medal)
- `medal` is `NA` when the athlete didn't win
- `noc` is the 3-letter country code (National Olympic Committee)
- Data covers both Summer and Winter Olympics

## 1.3 Clean and summarize

Our goal: **total medal counts by country for the 2016 Summer Olympics**.

This requires several steps chained with the pipe (`%>%`):

```{r}
#| warning: false

# Filter to Summer 2016 and only medal winners, then count by country
medals_by_country <- olympics_raw %>%
  filter(season == "Summer", year == 2016) %>%
  filter(!is.na(medal)) %>%
  group_by(noc) %>%
  summarize(
    total_medals = n(),
    gold = sum(medal == "Gold"),
    silver = sum(medal == "Silver"),
    bronze = sum(medal == "Bronze")
  ) %>%
  arrange(desc(total_medals))

# What did we get?
head(medals_by_country, 10)
```

Let's break down what each line did:

| Step | What it does |
|------|-------------|
| `filter(season == "Summer", year == 2016)` | Keep only Summer 2016 rows |
| `filter(!is.na(medal))` | Keep only medal winners |
| `group_by(noc)` | Group rows by country code |
| `summarize(...)` | Collapse each group into one row with counts |
| `arrange(desc(total_medals))` | Sort from most to fewest medals |

### Quick visualization

```{r}
#| warning: false

medals_by_country %>%
  slice_max(total_medals, n = 20) %>%
  ggplot(aes(x = reorder(noc, total_medals), y = total_medals)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(x = NULL, y = "Total Medals",
       title = "Top 20 Countries by Medal Count",
       subtitle = "2016 Summer Olympics") +
  theme_minimal()
```

### Exercise 1

> Create a version of `medals_by_country` that includes **all Summer Olympics from 2000 onward** (2000, 2004, 2008, 2012, 2016). Count the total medals per country across all five games. *Hint:* change the `filter()` condition.

```{r}
# Your code here


```


# Part 2: Load and Clean the World Bank Data

## 2.1 Read the data

```{r}
#| warning: false
#| message: false

# Read the World Bank data from the input folder
wb_raw <- read_csv("input/wb_indicators_2016.csv")

glimpse(wb_raw)
```

**What's in here?**

- `country`: Country name
- `iso2c` / `iso3c`: 2-letter and 3-letter country codes
- `gdp_pc`: GDP per capita in constant 2015 US dollars
- `population`: Total population

## 2.2 A problem: aggregates

The World Bank data includes **aggregate regions** (e.g., "Sub-Saharan Africa", "World") mixed in with actual countries. We need to remove these.

```{r}
#| warning: false

# How many rows before cleaning?
nrow(wb_raw)

# Look at some suspicious entries
wb_raw %>%
  filter(str_detect(country, "World|Africa|income|Euro"))  %>%
  select(country, iso2c, iso3c)
```

```{r}
#| warning: false

# Clean: remove aggregates and rows with missing GDP
wb_clean <- wb_raw %>%
  filter(!is.na(iso2c)) %>%
  filter(nchar(iso2c) == 2) %>%
  select(country, iso2c, iso3c, gdp_pc, population) %>%
  filter(!is.na(gdp_pc))

# How many actual countries remain?
nrow(wb_clean)
```

### Exercise 2

> What are the 10 richest countries by GDP per capita? The 10 poorest? Use `slice_max()` and `slice_min()`.

```{r}
# Your code here


```


# Part 3: Merging the Datasets

This is the key skill: **joining two datasets that don't share the same identifiers**.

## 3.1 The problem

The Olympics data uses `noc` (National Olympic Committee codes) and the World Bank uses `iso3c` (ISO 3-letter codes). They're *usually* the same... but not always.

Let's see which countries **don't match**:

```{r}
#| warning: false

# Which medal-winning countries have no match in the World Bank data?
unmatched <- medals_by_country %>%
  anti_join(wb_clean, by = c("noc" = "iso3c"))

unmatched
```

`anti_join()` shows rows in the first dataset with **no match** in the second. These are countries where the Olympic code differs from the ISO code (e.g., `GER` vs `DEU` for Germany).

## 3.2 Build a crosswalk

When codes don't match, we create a **crosswalk** — a small lookup table that maps one set of codes to another. This is a very common data wrangling task.

```{r}
#| warning: false

# Crosswalk: Olympic NOC code → ISO 3-letter code
noc_to_iso <- tribble(
  ~noc,  ~iso3c,
  "GER", "DEU",
  "DEN", "DNK",
  "SUI", "CHE",
  "NED", "NLD",
  "GRE", "GRC",
  "POR", "PRT",
  "CRO", "HRV",
  "RSA", "ZAF",
  "BUL", "BGR",
  "MAS", "MYS",
  "CHI", "CHL",
  "IRI", "IRN",
  "LAT", "LVA",
  "SIN", "SGP",
  "PHI", "PHL",
  "MGL", "MNG",
  "TPE", "TWN",
  "KOS", "XKX",
  "INA", "IDN",
  "VIE", "VNM",
  "BRN", "BHR",
  "PUR", "PRI",
  "FIJ", "FJI",
  "BAH", "BHS",
  "UAE", "ARE",
  "TGA", "TON",
  "KSA", "SAU",
  "NGR", "NGA"
)

# Apply the crosswalk: use ISO code if available, otherwise keep the NOC code
medals_by_country <- medals_by_country %>%
  left_join(noc_to_iso, by = "noc") %>%
  mutate(iso3c = ifelse(is.na(iso3c), noc, iso3c))
```

## 3.3 Merge

Now that we have a common key (`iso3c`), we can merge.

```{r}
#| warning: false

# Left join: keep all medal-winning countries, add World Bank data where available
merged <- medals_by_country %>%
  left_join(wb_clean, by = "iso3c")

# Did it work? Check how many matched
sum(!is.na(merged$gdp_pc))

# Which countries still didn't match?
merged %>% filter(is.na(gdp_pc)) %>% select(noc, iso3c)
```

### Types of joins

| Join | What it keeps |
|------|-------------|
| `left_join(a, b)` | All rows from `a`, matching rows from `b` |
| `right_join(a, b)` | All rows from `b`, matching rows from `a` |
| `inner_join(a, b)` | Only rows that match in **both** |
| `full_join(a, b)` | All rows from **both** |
| `anti_join(a, b)` | Rows in `a` with **no match** in `b` |

### Exercise 3

> Use `inner_join()` instead of `left_join()` to merge the data. How many rows do you get? Why is it different?

```{r}
# Your code here


```

## 3.4 Save your cleaned, merged data

Always save your cleaned data to the `output/` folder — never overwrite your raw data in `input/`.

```{r}
# Save the merged dataset
write_csv(merged, "output/olympics_wb_merged.csv")
```


# Part 4: Analysis

Now we can ask: **Does national wealth predict Olympic success?**

## 4.1 Create new variables

```{r}
#| warning: false

merged <- merged %>%
  filter(!is.na(gdp_pc), !is.na(population)) %>%
  mutate(
    medals_per_million = total_medals / (population / 1e6),
    log_gdp_pc = log(gdp_pc),
    log_pop = log(population)
  )
```

## 4.2 Visualize

```{r}
#| warning: false

ggplot(merged, aes(x = log_gdp_pc, y = total_medals)) +
  geom_point(aes(size = population / 1e6), alpha = 0.6, color = "steelblue") +
  geom_text(
    data = merged %>% slice_max(total_medals, n = 8),
    aes(label = noc), nudge_y = 5, size = 3
  ) +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linewidth = 0.8) +
  scale_size_continuous(name = "Population (millions)", range = c(1, 12)) +
  labs(
    x = "Log GDP per Capita (constant 2015 US$)",
    y = "Total Medals",
    title = "Does National Wealth Predict Olympic Success?",
    subtitle = "2016 Summer Olympics"
  ) +
  theme_minimal()
```

```{r}
#| warning: false

ggplot(merged, aes(x = log_gdp_pc, y = medals_per_million)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_text(
    data = merged %>% slice_max(medals_per_million, n = 8),
    aes(label = noc), nudge_y = 0.5, size = 3
  ) +
  geom_smooth(method = "lm", se = TRUE, color = "firebrick", linewidth = 0.8) +
  labs(
    x = "Log GDP per Capita (constant 2015 US$)",
    y = "Medals per Million People",
    title = "Olympic Success per Capita vs. National Wealth",
    subtitle = "2016 Summer Olympics"
  ) +
  theme_minimal()
```

## 4.3 Regression

```{r}
#| warning: false

models <- list()
models[["Bivariate"]] <- lm(total_medals ~ log_gdp_pc, data = merged)
models[["+ Population"]] <- lm(total_medals ~ log_gdp_pc + log_pop, data = merged)
models[["Per Capita"]] <- lm(medals_per_million ~ log_gdp_pc, data = merged)
models[["Per Capita + Pop"]] <- lm(medals_per_million ~ log_gdp_pc + log_pop, data = merged)

modelsummary(
  models,
  estimate  = "{estimate}{stars} ({std.error})",
  statistic = NULL,
  gof_omit = "IC|RMSE|Log|F|R2$|Std."
)
```

### Exercise 4

> Interpret the coefficient on `log_gdp_pc` in the "+ Population" model. What does it tell us about the relationship between wealth and Olympic medals, controlling for population size?

### Exercise 5

> Why does the story change when we look at medals *per capita* vs. total medals? Which do you think is a better measure of Olympic "success"?


# Part 5: Extensions (if time permits)

## 5.1 Multiple years

> **Challenge:** Create medal counts for multiple Olympic years (2000-2016). You'll need to download World Bank data for matching years too using the `WDI` package:

```{r}
#| eval: false

# install.packages("WDI")  # run this once if you don't have it
library(WDI)

wb_panel <- WDI(
  indicator = c(gdp_pc = "NY.GDP.PCAP.KD", population = "SP.POP.TOTL"),
  country = "all",
  start = 2000,
  end = 2016
)
```

```{r}
# Your code here


```

## 5.2 Add another indicator

> **Challenge:** Add an additional World Bank indicator and see if it improves your model. Some useful codes:

| Code | Indicator |
|------|-----------|
| `SE.XPD.TOTL.GD.ZS` | Government expenditure on education (% of GDP) |
| `SI.POV.GINI` | Gini index (inequality) |
| `SH.XPD.CHEX.GD.ZS` | Health expenditure (% of GDP) |
| `SL.UEM.TOTL.ZS` | Unemployment rate |

```{r}
# Your code here


```


# Summary of Key Functions

| Function | What it does |
|----------|-------------|
| `read_csv()` | Import CSV from a file or URL |
| `glimpse()` | Quick look at data structure |
| `filter()` | Keep rows that meet a condition |
| `select()` | Keep or rename columns |
| `mutate()` | Create or modify columns |
| `group_by() + summarize()` | Aggregate data by groups |
| `left_join()` | Merge two datasets by a shared key |
| `anti_join()` | Find rows with no match |
| `tribble()` | Create small lookup tables by hand |
| `write_csv()` | Save data to a CSV file |
| `slice_max() / slice_min()` | Keep top/bottom rows |
| `arrange()` | Sort rows |
